Boutiques: a flexible framework for automated application integration in computing platforms

### 2016-11-10 paper.tex
Proof-read sections I and II

\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\todo}[1]{\color{red}TODO: #1\color{black}}
\newcommand{\boutiques}{Boutiques\xspace}
\algrenewcommand{\algorithmiccomment}[1]{\# \textit{#1}} % so that comments in algorithms are preprended by '#' instead of a small right arrow

@@

\abstract{Porting applications to execution platforms is a critical
  step to enable their sharing among scientific communities and their
  efficient execution on computing infrastructures. However,
  application porting remains a costly human effort that consists of
  1)~installing the application on the target infrastructure,
  2)~describing the application in a format compatible with the
  execution platform, and 3)~generating proper user interfaces. Due to
  the variety of platforms, application porting efforts are often
  replicated several times while mutualization would save cost and
  improve the quality of the ported applications. We describe
  \emph{boutiques}, a system that allows for automatic import and
  exchange of applications across platforms. Boutiques relies on Linux
  containers to solve the problem of application installation in a
  lightweight manner. It also adopts a rich and flexible application
  description format which is versatile enough to be used in various
  platforms. Boutiques is currently supported by \todo{X} platforms
  and was used to describe \todo{Y} applications.
}

@@

Platforms such as web services, portals, virtual research
environments, science gateways and workflow engines often rely on
third-party data processing and simulation applications to deliver
services to their users. Such applications are often repeatedly and
manually ported to various platforms whereas this porting effort would
greatly benefit from being mutualized. Virtual container systems such
as Docker and Singularity greatly facilitate the sharing of
applications executed in Linux systems, and improve the
reproducibility of analyses by defining immutable, reusable execution
environments. However, containers need to be properly annotated to
allow platforms to leverage the applications that they
contain. Basically, the command lines that can be run in a container
must be formally described so that they can be automatically created
and executed by platforms. Describing command lines is not
straightforward since they often contain implicit semantics specified
in human-readable documents such as manuals.

In Section~\ref{sec:system}, we describe \boutiques, a system to share
command-line applications across platforms. In \boutiques, a command
line is described from a flexible template referring to the inputs it
requires and the outputs it produces. Inputs may be inter-dependent,
e.g., input A is available only if input B was defined and passed in
configuration files or directly on the command line. Such formal
descriptions, also called \emph{manifests}, allows for advanced
validation of input values, which is useful error detection. Manifests
refer to containers where applications and all their dependencies are
installed. \boutiques also provides a simple parallelization framework
that allows applications such as workflow engines to execute multiple
tasks concurrently. A set of core tools facilitates the construction,
validation and execution of \boutiques manifests. A repository
orchestrates these tools together.

\boutiques was created using a bottom-up approach in which new
features were progressively added as applications and platforms
required. It was used to describe \todo{Y} applications and it is
supported by \todo{X} platforms, which we report in
Section~\ref{sec:results}.

\boutiques manifests are intented to be produced by scientific
application developers, to be published in common repositories, and to
be consumed by execution platforms. Although the motivating use cases
were taken from neuroinformatics, our primary field of interest,
nothing prevents the system from being used in a variety of other
fields.

\note{Tristan}{Vocabulary (check consistency across the paper): tool vs. application; descriptor vs. manifest; task vs. invocation}
\note{Tristan}{Targeted journals: Scientific Data, Giga Science, Frontiers in Neuroinformatics, Future Generation Computer Systems}

@@

The application is described with a \emph{manifest} that specifies the
command-line template, inputs and outputs. The manifest may point to a
container where the application and all its dependencies are
installed. It may also contain an invocation schema used for input
validation. At runtime, the execution \emph{platform} builds the
application command line from the manifest and values entered by the
user. The platform runs the command line on the execution
infrastructure, e.g., a server, a cluster or a cloud, within a
container whenever available. To build and run the command line, the
platform may rely on the Boutiques \emph{core tools}, in particular on
the Boutiques validator and local executor. Application manifests may
be stored in the \emph{repository}, which provides additional features.

@@

The core component of the JSON manifest is a flexible command-line
description represented as a simple string template complying to the
syntax of the \texttt{bash} Linux shell interpreter. The command line
template may contain placeholders for input and output values, called
value keys. The value keys may be any string, and the
command line may actually encompass several commands, e.g., separated
by semicolumns, pipes (\texttt{|}) or ampersands (\texttt{\&}). Such a
flexible command-line specification is meant to facilitate the
embedding of minor operations on the command line, for instance input
decompression and output archival, avoiding the use of multiple
``shims'' to align mismatching applications in a
pipeline~\cite{hull2004treating}.

Here is an example of a typical command-line template:
\begin{verbatim}
exampleTool_1 [STRING_INPUT] [FILE_INPUT] [ENUM_INPUT] | \
                              exampleTool_2 [FLAG_INPUT] [NUMBER_INPUT] >> [LOG].txt
\end{verbatim}
The template contains six value keys for input and output
values. When the application is executed, such keys will be replaced
by values and file names according to the user input. Flags will also
be added wherever appropriate, with customizable separators. No
particular syntax is imposed for the value keys, they just have
to be unique.  Note the use of the pipe (\texttt{|}) operator to chain
tools, and of the \texttt{>>} operator to redirect the standard output
to a file.

@@

\paragraph{General properties.} Inputs must have a name, a unique
identifier and a type. They may be optional, have a description, a
value key, a flag and flag separator, and a default
value. Inputs may also be ordered lists; in this case, value
keys are substituted by the space-separated list of input values.

\paragraph{Types.} Inputs may be of type \texttt{String},
\texttt{Number}, \texttt{Flag} or \texttt{File}. \texttt{File} may
also represent a directory. Types can be restricted to a specific set
of values (\texttt{String} and \texttt{Number} only), to a specific
range (\texttt{Number} only), to integers or to booleans (\texttt{Number}
only).

\paragraph{Groups and dependencies.} Groups of inputs may be defined
from a identifier, name and list of input identifiers. Groups may be
used to improve the presentation in a graphical user interface and to
specify the following constraints among inputs: (1)
\texttt{mutually-exclusive}: only one input in the group may have a
value; (2) \texttt{one-is-required}: at least one input in the group
must have a value. Dependencies among inputs may also be defined
regardless of a particular group: an input may (1) require a list of
inputs (2) disable a list of inputs.

@@

Application outputs are the files or directories that need to be
transferred and delivered to the user once the execution is
complete. Although it may not be required to distinguish inputs from
outputs to build the application command line, this information is
needed by computing platforms to identify the files that must be
saved after the execution.

In Boutiques, output files are defined from a unique identifier, a
name and a path template that specifies the file or directory
name. Path templates may include input value keys in case
output files are named after the input values. Input values may be
stripped from specific strings, e.g., file extensions before being
substituted in the path template. Output files may also have a
description, a command-line flag, a flag separator, and a value
key in case they appear on the command line. They may be optional in
case the file is not always produced by the application, for instance
when it is produced only when a particular flag is activated. They
may also be (un-ordered) lists; in this case, the path template must
contain a wildcard (\texttt{*}) matching any string of characters
and defining the pattern used to match the
output files in the list.

@@

{
    "id" : "logfile",
    "name" : "Log file",
    "description" : "The output log file from the example tool",
    "path-template" : "log-[STRING_INPUT]",
    "value-key" : "[LOG]",
    "path-template-stripped-extensions" : [".txt", ".csv"],
    "optional" : false
}
\end{minted}
\caption{Output file example} 
\label{listing:output-example}
\end{listing}

@@

Configuration files may be complex though, and specified in any
language.  As it does not seem reasonable to adopt a particular
template for configuration files, the Boutiques specification allows
application developers to specify their own template, referring to
input and output value keys. Configuration files are specific types of
output files that have a
 file template defined. As output files, 
they must have a path template that defines
how they must be named and where they must be written. They may also
have a value key and a flag in case they need to be passed on the
command line.

Listing~\ref{listing:configuration-file-example} shows an example.
\begin{listing}
\begin{minted}[frame=single,
               framesep=3mm,
               linenos=false,
               xleftmargin=21pt,
               tabsize=4]{js}
               {
                   "id": "config_file",
                   "name": "Configuration file",
                   "type": "Configuration File",
                   "value-key": "[CONFIG_FILE]",
                   "path-template": "config.txt",
                   "file-template": [
                   "# This input is hard-coded",
                   "stringInput=foo",
                   "fileInput=[FILE-INPUT]",
                   "# And here is the result",
                   "fileOutput=[OUTPUT-FILE]",
                   ""
                   ]
                 }
\end{minted}
\caption{Configuration file example. The file template is defined as
  an array of strings to allow for multi-line strings in JSON.}
\label{listing:configuration-file-example}
\end{listing}

Rigorous input validation is an important motivation for
Boutiques. The obvious solution in our context is to rely on an
application-specific JSON schema, called ``invocation schema'', to
specify the correct input values that an application
accepts. Platforms could rely on the invocation schema to validate
inputs automatically using any JSON validator, without having to
develop specific code.

Invocation schemas, however, are complex JSON objects. Basically, they
must represent the properties described above in a formal way,
including dependencies between
inputs. Listing~\ref{listing:invocation-schema-example} shows an
example of how dependencies between mutually exclusive parameters are
defined in the invocation schema.

The invocation schema also has a \texttt{task-dependencies} array that
specifies that the new invocation to start only when all the
invocations in this array have completed. This property is used in the
parallelization model described hereafter.

It does not seem realistic to assume that developers will easily write
correct JSON schemas for their applications. Instead, invocation
schemas should be generated automatically using a core Boutiques
tool. Accordingly, invocation schema is an optional property of the
Boutiques manifest, that platforms or developers may generate using
the appropriate tool.

\subsection{Parallelization support}
\label{sec:parallelization}
Boutiques support parallelism by allowing applications to submit new
tasks through the following protocol:
\begin{enumerate}
\item Applications write a JSON object complying to the invocation
  schema in the directory where there were launched.
\item The platform reads the JSON object, creates the corresponding
  task, and writes back the task id in the same directory.
% The platform could also support status requests.
\end{enumerate}
Several invocations can be submitted at once, possibly using the
\texttt{task-dependencies} array to specify dependencies among
them. This mechanism is implemented in the Boutiques schema using a
single property, \texttt{can-submit-tasks}.

Although very simple, this model allows wrapping complete
workflows. Workflows are wrapped as any other application, and they
can be executed using any engine such as Nipype, PSOM and others,
provided that such engines are installed in the container image
referred in the application manifest.

The adopted ``task encapsulation'' model is very simple but it
allows for a scalable and reliable execution of workflows expressed in
a variety of languages, as detailed in ~\cite{glatard2016fgcs}.

The implementation of the application is defined in a container image
in the Docker, Singularity or rootfs format. We intentionally support
multiple container formats as we anticipate that they will be used for
different purposes. For instance, Docker is well suited for
application developers and users who want to use applications on their
local workstation. It is very well documented, maintained and it has a
rich ecosystem of tools to help build and run containers on most
platforms, including Windows, MacOS and Linux. Singularity is more
suited for users and platforms that need to run applications on shared
computing clusters. Bridges exist among these containers formats to
convert container images across frameworks. For instance, a platform
dedicated to high-performance computing may accept manifests referring
to Docker containers to facilitate application integration by
developers, and internally convert container images to Singularity to
run applications efficiently on clusters.

Containers are adopted because they allow for an automated and
lightweight integration of application implementations in
platforms. They are extremely useful to improve the reproducibility of
analyses as variations in the software environment may have an
important impact on the computed results. They also have limitations,
in particular they do not specify the hardware architecture required
to execute an application, which can be an issue in some cases.

Custom properties may be added to the Boutiques specification without
restriction. They may be useful to implement platform-specific
features but they should be used with care to avoid making tools
dependent on a particular platform.

@@

At runtime, a value is assigned to all the mandatory and some of the
optional inputs. Inputs of type "String" may contain any string,
inputs of type "Number" must contain a string representing a number,
inputs of type "File" must contain a string representing a file path
(absolute or relative to the execution directory) and inputs of type
"Flag" must contain a boolean. When input is a list, the value
contains the concatenation of all strings in the list.

Algorithm~\ref{algo:command-line} shows how the application command
line is constructed from the application manifest and values entered
by the user. It substitutes all the value keys in the command line,
output path templates and configuration files, and writes the
configuration files.
\begin{algorithm}[h!]
\caption{Command-line construction}
\label{algo:command-line}
\begin{algorithmic}
  \State \Comment{Substitute input value keys in output path templates, configuration files and command line.}
  \For{\texttt{input} in \texttt{inputs}}
  \If{\texttt{input} has a \texttt{value-key}}
  \For{\texttt{output} in \texttt{outputs}}
  \State \texttt{stripped\_value} = \texttt{input\_value}
  \If{\texttt{input} type is \texttt{File}}
  \State In \texttt{stripped\_value}, remove  all elements in \texttt{path-template-stripped-extensions}.
  \EndIf
  \State In \texttt{path-template}, replace all occurrences of \texttt{value-key} by \texttt{stripped\_value}.
  \If{\texttt{output} has a \texttt{file-template}}
  \State In any line of \texttt{file-template}, replace all occurrences of \texttt{value-key} by \texttt{stripped\_value}.
  \EndIf
  \EndFor
  \State Prepend \texttt{command-line-flag} and \texttt{command-line-flag-separator} to \texttt{input\_value}.
  \State In \texttt{command-line}, replace all occurrences of \texttt{value-key} by \texttt{input\_value}.
  \EndIf
  \EndFor

@@

@@

The Boutiques validator checks conformance of JSON manifests to the
Boutiques schema using a basic JSON validator. It also performs an
additional set of checks that cannot be easily implemented in JSON
schema: value keys are unique among inputs, input and output
identifiers are unique, input and output value keys are all
included in the command line, value keys are not contained
within each other (which would puzzle substitution), output path
templates are unique (to avoid results to overwrite each other),
inputs of type Flag have a command-line flag, are optional and are not
lists, number and list constraints are sensible (e.g. min is lower
than max), the default value of enum types is part of the enum, an
input cannot both require and disable another input, required inputs
cannot require or disable other parameters, group member identifiers
must correspond to existing inputs and cannot appear in different
groups, mutually exclusive groups cannot have members requiring other
members, and that one-is-required groups should never have required
members. The Boutiques validator is meant for application developers.

The local executor can generate and execute command line from a
Boutiques manifest and a set of input values represented in a CSV or
JSON file complying with the invocation schema. It runs the
command in a container provided that the required framework
(e.g. Docker) is already installed. It can also generate hypothetical
command lines from random values, for debugging purposes. The local
executor can be used by application users, to run tools locally, or by
platforms, to generate command lines to be run on the execution
infrastructure.

The invocation schema generator creates an invocation schema (see
Section~\ref{sec:invocation-schema}) from a Boutiques manifest and
validates data against it. It can be used by platform administrators
to add invocation schemas to existing manifests, and by platforms to
validate input data. Validation is done using a basic JSON
validator.

\todo{json editor}

@@

CBRAIN~\cite{SHER-14} is a web platform that can process data
distributed into multiple storage locations on computing clusters and
clouds. CBRAIN offers transparent access to remote data sources,
distributed computing sites, and an array of processing and
visualization tools within a controlled, secure environment.  The
CBRAIN service deployed at the Montreal Neurological Institute relies
on the infrastructure provided by Compute Canada~\cite{das2016mni}. It
currently provides 460+ collaborators in 20 countries with web access
to several systems, including 6 clusters of the Compute Canada
high-performance computing infrastructure (totalling more than 100,000
computing cores and 40PB of disk storage) and Amazon EC2. CBRAIN
transiently stores about 10 million files representing over 50TB
distributed in 42 servers. 56 data processing tools are integrated and
over 340,000 processing batches have been submitted since 2010.

@@

\paragraph{Niak.} The Niak pre-processing pipeline, executed with the Pipeline System
for Octave and Matlab (PSOM)~\cite{bellec2012pipeline}, was integrated
in CBRAIN using Boutiques. The integration uses the parallelization
mechanism described in Section~\ref{sec:parallelization} so that even
the tasks processing a single subject can be parallelized. It also
allows CBRAIN to leverage the efficient agent model used in PSOM, as
described in~\cite{GLAT-16}. The integration required some work in
PSOM to facilitate its invocation as a non-interactive command-line
application. The resulting CBRAIN plugin is available at
\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It could in
principle be used in other platforms that support the parallelization
property, although this has not been tested.

@@

\paragraph{Other tools.} Other neuroinformatics tools were ported to CBRAIN using Boutiques:
FSL tools and ICA AROMA are available at
\url{https://github.com/aces/cbrain-plugins-neuro} and some pipelines
of the Human Connectome Project (HCP) are at
\url{https://github.com/big-data-lab-team/cbrain-plugins-hcp}.

@@

\begin{itemize}
\item Security: containers may be updated after being integrated in the platforms.
\item The manifest specification is getting increasingly complex, but
  this complexity is required to describe real-world
  applications. When something cannot be described using the manifest,
  it is always possible to change the command line format by writing a
  wrapper script for the application. However, when containers are
  used, the wrapper script needs to be added to the container, which
  has some overhead (in particular, the containers provided by the
  developers cannot be used directly).
\item Testing is critical.
\item Workflow engines can be platforms (they can run Boutiques tools), workflows can be applications (they can be described in Boutiques).
\item With Boutiques, developers can port their application once and execute it in several platforms. It removes the technological dependency to a particular platform and facilitates migration. 
\end{itemize}

@@

Boutiques intentionally does not support the definition of complete
workflows from individual applications. Indeed, we believe that
workflow engines are specific tools that need complex languages to
implement the particular functions that they target. Efforts to
specify common workflow languages include the Common Workflow Language
(CWL)~\cite{cwl} and the Interoperable Workflow Intermediate
Representation (IWIR)~\cite{plankensteiner2011iwir}.

>>>>

\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\todo}[1]{\color{red}TODO: #1\color{black}}
\newcommand{\boutiques}{Boutiques\xspace}
\newcommand{\notimplementedyet}[1]{\color{blue}\emph{#1}\footnote{Still needs to be implemented}\color{black}\xspace}
\algrenewcommand{\algorithmiccomment}[1]{\# \textit{#1}} % so that comments in algorithms are preprended by '#' instead of a small right arrow

@@

\abstract{Porting applications to execution platforms such as web
  portals and workflow engines is critical to enable their sharing
  among scientific communities and their efficient execution on
  computing infrastructures. However, application porting remains a
  costly human effort that consists of 1)~installing the application
  on the target infrastructure, 2)~describing the application in a
  format compatible with the execution platform, and 3)~generating
  proper user interfaces. Due to the variety of platforms, application
  porting efforts are often replicated several times while
  mutualization would save cost and improve the quality of the ported
  applications. We describe \boutiques, a system that allows for
  automatic import and exchange of applications across
  platforms. \boutiques adopts a rich, flexible application
  description format and it relies on Linux containers to install
  applications across platforms in a lightweight manner. It is
  currently supported by \todo{X} platforms and was used to describe
  \todo{Y} applications.  }

@@

Platforms such as web services, portals, virtual research
environments, science gateways and workflow engines often rely on
third-party data processing and simulation applications to deliver
services to their users. Such applications are often repeatedly and
manually ported to various platforms whereas this porting effort would
greatly benefit from being mutualized. Meanwhile, virtual container
systems such as Docker and Singularity greatly facilitate the sharing
of software and improve the reproducibility of analyses by defining
immutable, reusable execution environments. However, containers need
to be properly annotated to allow platforms to leverage the software
they contain. Basically, the command lines that can be run in a
container must be formally described so that they can be automatically
created and executed by platforms. Describing command lines is not
straightforward since they often contain implicit semantics specified
in human-readable documents such as manuals.

\boutiques is a system to share command-line applications across
platforms. In \boutiques, a command line is described using a flexible
template comprising the inputs it requires and the outputs it
produces. Inputs may be inter-dependent, e.g., input A cannot be used
if input B is used, and passed in configuration files or directly on
the command line. Such formal descriptions, also called
\emph{manifests}, allow for advanced validation of input values to
prevent errors. Manifests refer to containers where applications and
all their dependencies are installed. \boutiques also provides a
simple parallelization framework that allows applications such as
workflow engines to concurrently execute multiple tasks (in the
remainder, tasks are called \emph{invocations}). A set of core tools
facilitate the construction, validation and execution of \boutiques
manifests. A repository orchestrates these tools together. \boutiques
manifests are intented to be produced by scientific application
developers, to be published in common repositories and to be consumed
by execution platforms.

The remainder of this paper describes the \boutiques system
(Section~\ref{sec:system}) and reports on its adoption by platforms
and applications in the neuroinformatics domain, our primary field of
interest (Section~\ref{sec:results}). It closes on a discussion and
comparison with related systems.

\note{Tristan}{Vocabulary (check consistency across the paper): tool vs. application}
\note{Tristan}{Targeted journals: Scientific Data, Giga Science, Frontiers in Neuroinformatics, Future Generation Computer Systems}

@@

Applications are described with a JSON manifest that specifies the
command-line template, inputs and outputs. The manifest may point to a
container where the application and all its dependencies are
installed. It may also contain an invocation schema used for input
validation. At runtime, the execution platform builds the command line
from the manifest and values entered by the user. The platform runs
the command line on the execution infrastructure, e.g., a server, a
cluster or a cloud, within a container whenever available. To build
and run the command line, the platform may rely on the \boutiques core
tools, in particular on the validator and local executor. Application
manifests may be stored in the repository which provides additional
features.

@@

The core component of the manifest is a flexible command-line
description represented as a simple string template complying to the
syntax of the \texttt{bash} UNIX shell. \texttt{bash} is the default
shell on the major Linux distributions and OS X. The command-line
template may contain placeholders for input and output values, called
value keys. The command line may actually
encompass several commands separated, e.g., by semicolumns, pipes
(\texttt{|}) or ampersands (\texttt{\&}). Such a flexible command-line
specification is meant to facilitate the embedding of minor mundane
operations on the command line, for instance directory creation, input
decompression or output archival.

Here is an example of a typical command-line template:
\begin{verbatim}
exampleTool_1 [STRING_INPUT] [FILE_INPUT] [ENUM_INPUT] | \
                              exampleTool_2 [FLAG_INPUT] [NUMBER_INPUT] >> [LOG].txt
\end{verbatim}
The template contains six value keys for input and output values, that
will be replaced by values and file names according to the user input
when the application is executed. Flags will also be added wherever
appropriate, with customizable separators. No particular syntax is
imposed for the value keys, they just have to be unique.  Note the use
of the pipe (\texttt{|}) operator to chain applications, and of the
\texttt{>>} operator to redirect the standard output to a file.

@@

\paragraph{General properties.} Inputs must have a name, a unique
identifier and a type. They may be optional, have a description, a
value key, a flag and flag separator, and a default value. Inputs may
also be ordered lists: in this case, value keys are substituted by the
space-separated list of input values.

\paragraph{Types.} Inputs may be of type \texttt{String},
\texttt{Number}, \texttt{Flag} or \texttt{File}. \texttt{File} may
also represents a directory. Types can be restricted to a specific set
of values (\texttt{String} and \texttt{Number} only), to a specific
range (\texttt{Number} only), to integers or to booleans (\texttt{Number}
only).

\paragraph{Groups and dependencies.} Groups of inputs may be defined
with a identifier, name and list of input identifiers. Groups may be
used to improve the presentation in a graphical user interface and to
specify the following constraints among inputs: (1)
\texttt{mutually-exclusive}: only one input in the group may have a
value; (2) \texttt{one-is-required}: at least one input in the group
must have a value. Dependencies among inputs may also be defined
regardless of a particular group: an input may (1) require a list of
inputs and (2) disable a list of inputs.

@@

Application outputs are files and directories that need to be
transferred and delivered to the user once the execution is
complete. Although it may not be required to distinguish inputs from
outputs to build the application command line, this information is
needed by computing platforms to identify the files that must be saved
after the execution.

In \boutiques, output files must have a unique identifier, a name and
a path template that specifies the file or directory name. Path
templates may include input value keys in case output files are named
after the input values. In this case, input values may be stripped
from specific strings, e.g., file extensions before being substituted
in the path template. Output files may also have a description, a
command-line flag, a flag separator, and a value key in case they
appear on the command line. They may be optional in case the file is
not always produced by the application, for instance when it is
produced only when a particular flag is activated. They may also be
(un-ordered) lists: in this case, the path template must contain a
wildcard (\texttt{*}) matching any string of characters and defining
the pattern used to match the output files in the list.

@@

{
    "id" : "logfile",
    "name" : "Log file",
    "description" : "The output log file from the example application",
    "path-template" : "log-[STRING_INPUT]",
    "value-key" : "[LOG]",
    "path-template-stripped-extensions" : [".txt", ".csv"],
    "optional" : false
}
\end{minted}
\caption{Output file example} 
\label{listing:output-example}
\end{listing}

@@

Configuration files may be complex though, and specified in any
language.  As it does not seem reasonable to adopt a particular
template for configuration files, the \boutiques specification allows
application developers to specify their own template containing input
and output value keys. Configuration files are specific types of
output files that must have a file template in addition to a path
template that defines how they will be named and where they will be
written. They may also have a value key and a flag in case they need
to be passed on the command
line. Listing~\ref{listing:configuration-file-example} shows an
example.
\begin{listing}
\begin{minted}[frame=single,
               framesep=3mm,
               linenos=false,
               xleftmargin=21pt,
               tabsize=4]{js}
               {
                   "id": "config_file",
                   "name": "Configuration file",
                   "type": "Configuration File",
                   "value-key": "[CONFIG_FILE]",
                   "path-template": "config.txt",
                   "file-template": [
                   "# This input is hard-coded",
                   "stringInput=foo",
                   "fileInput=[FILE-INPUT]",
                   "# And here is the result",
                   "fileOutput=[OUTPUT-FILE]",
                   ""
                   ]
                 }
\end{minted}
\caption{Configuration file example. The file template is defined as
  an array of strings to allow for multi-line strings in JSON.}
\label{listing:configuration-file-example}
\end{listing}

@@

At runtime, a value is assigned to all the mandatory and some of the
optional inputs. %% Inputs of type ``String'' may contain any string,
%% inputs of type "Number" must contain a string representing a number,
%% inputs of type "File" must contain a string representing a file path
%% (absolute or relative to the execution directory) and inputs of type
%% "Flag" must contain a boolean. 
%% When input is a list, the value contains the concatenation of all the
%%values in the list.
Algorithm~\ref{algo:command-line} shows how the command line is
constructed from the manifest and the values entered by the user. It
substitutes all the value keys in the command line, output path
templates and configuration files, and writes the configuration files.
\begin{algorithm}[h!]
\caption{Command-line construction}
\label{algo:command-line}
\begin{algorithmic}
  \State \Comment{Substitute input value keys in output path templates, configuration files and command line.}
  \For{\texttt{input} in \texttt{inputs}}
  \If{\texttt{input} has a \texttt{value-key}}
  \For{\texttt{output} in \texttt{outputs}}
  \State \texttt{stripped\_value} = \texttt{input\_value}
  \If{\texttt{input} type is \texttt{File}}
  \State In \texttt{stripped\_value}, remove  all elements in \texttt{path-template-stripped-extensions}.
  \EndIf
  \State In \texttt{path-template}, replace all occurrences of \texttt{value-key} by \texttt{stripped\_value}.
  \If{\texttt{output} has a \texttt{file-template}}
  \State In any line of \texttt{file-template}, replace all occurrences of \texttt{value-key} by \texttt{stripped\_value}.
  \EndIf
  \EndFor
  \State Prepend \texttt{command-line-flag} and \texttt{command-line-flag-separator} to \texttt{input\_value}.
  \State In \texttt{command-line}, replace all occurrences of \texttt{value-key} by \texttt{input\_value}.
  \EndIf
  \EndFor

@@

Rigorous input validation is an important motivation for
\boutiques. The obvious solution for JSON documents is to rely on an
application-specific JSON schema, called \emph{invocation schema}, to
specify the input values accepted by an application. Platforms can
rely on invocation schemas to validate inputs using
any JSON validator, without having to develop specific code.

Invocation schemas, however, are complex JSON objects. Basically, they
must represent the properties described above in a formal way,
including dependencies between
inputs. Listing~\ref{listing:invocation-schema-example} shows an
example of how dependencies between mutually exclusive parameters are
defined in the invocation schema. As it does not seem realistic to
assume that developers will easily write correct invocations schemas
even for their own applications, invocation schemas can be generated
automatically using a core \boutiques tool. The invocation schema is
stored as an optional property of the Boutiques manifest.

\notimplementedyet{The invocation schema also has a
  \texttt{task-dependencies} array that allows specifying that the new
  invocation may start only when previously-submitted invocations have
  completed. This property is used in the parallelization model
  described hereafter.}

\subsection{Parallelization support}
\label{sec:parallelization}
\boutiques supports parallelism by allowing applications to submit new
invocations through the following protocol \todo{Needs to be
  documented on Github}:
\begin{enumerate}
\item Applications write a JSON object complying to the invocation
  schema in their execution directory.
\item The platform reads the JSON object, creates the corresponding
  invocation, and writes back the invocation id in the same directory.
% The platform could also support status requests.
\end{enumerate}
Several invocations can be submitted at once, possibly using the
\texttt{task-dependencies} array to specify dependencies among
them. This mechanism is implemented in the Boutiques schema using a
single property, \texttt{can-submit-tasks}.

Although very simple, this model allows wrapping complete
workflows. Workflows are wrapped as any other application, and they
can be executed using any engine provided that it is installed in the
container image referred in the application manifest. Such a ``task
encapsulation'' model allows for a scalable and reliable execution of
workflows expressed in a variety of languages, as detailed in
~\cite{glatard2016fgcs}.

Applications are installed in a container image complying to the
Docker, Singularity or rootfs format. We intentionally support
multiple container formats as we anticipate that they will be used for
different purposes. For instance, Docker is well suited for developers
and users who manipulate applications on their local workstation. It
is well documented, maintained and it has a rich ecosystem of tools to
help build and run containers on most operating systems. Singularity
is more suited for users and platforms that need to run applications
on shared computing clusters. Bridges exist among these containers
formats to convert container images across frameworks. For instance, a
platform dedicated to high-performance computing may accept manifests
referring to Docker containers to facilitate application integration
by developers, and internally convert container images to Singularity
to run applications efficiently on clusters.

Containers were adopted because they allow for an automated and
lightweight integration of application implementations in
platforms. They are extremely useful to improve the reproducibility of
analyses as variations in the software environment may have an
important impact on the computed results. They also have limitations,
in particular they do not specify the hardware architecture required
to execute an application, which can be an issue in some cases.

Custom properties may be added to the Boutiques specification without
restriction. Custom properties are corralled in a specific JSON object
to facilitate validation. They may be useful to implement
platform-specific features but they should be used with care to avoid
making applications dependent on a particular platform.

@@

The Boutiques validator checks conformance of JSON manifests to the
Boutiques schema using a basic JSON validator. It also performs the
following checks that cannot be easily implemented in JSON schema:
value keys are unique among inputs, input and output identifiers are
unique, input and output value keys are all included in the command
line, value keys are not contained within each other (which would
puzzle substitution), output path templates are unique (to avoid
results to overwrite each other), inputs of type Flag have a
command-line flag, are optional and are not lists, number and list
constraints are sensible (e.g. min is lower than max), the default
value of enum types is part of the enum, an input cannot both require
and disable another input, required inputs cannot require or disable
other parameters, group member identifiers must correspond to existing
inputs and cannot appear in different groups, mutually exclusive
groups cannot have members requiring other members, and
one-is-required groups should never have required members. The
Boutiques validator is meant for application developers; platforms
should rely on regular JSON validation to avoid depending on a
third-party tool.

The local executor can generate and execute command line from a
Boutiques manifest and a set of input values represented in a CSV or
JSON file complying with the invocation schema. It runs the command in
a container provided that the required framework (e.g. Docker) is
already installed. It can also generate hypothetical command lines
from random values for debugging purposes. The local executor can be
used by application users to run applications locally, or by platforms
to generate command lines to be run on the execution infrastructure.

The invocation schema generator can create an invocation schema (see
Section~\ref{sec:invocation-schema}) from a \boutiques manifest and
validate input data against it using a regular JSON validator. It can
be used to add invocation schemas to existing manifests.

Finally, a JSON editor facilitates the writing of \boutiques manifests
through a web interface.

@@

CBRAIN~\cite{SHER-14} is a web platform that can process data
distributed into multiple storage locations on computing clusters and
clouds. CBRAIN offers transparent access to remote data sources,
distributed computing sites, and an array of processing and
visualization tools within a controlled, secure environment.  The
CBRAIN service deployed at the Montreal Neurological Institute relies
on the infrastructure provided by Compute Canada~\cite{das2016mni}. It
currently provides 460+ collaborators in 20 countries with web access
to several systems, including 6 clusters of the Compute Canada
high-performance computing infrastructure (totalling more than 100,000
computing cores and 40PB of disk storage) and Amazon EC2. CBRAIN
transiently stores about 10 million files representing over 50TB
distributed in 42 servers. 56 data processing applications are integrated and
over 340,000 processing batches have been submitted since 2010.

@@

\paragraph{Niak.} The Niak pre-processing pipeline, executed with the Pipeline System
for Octave and Matlab (PSOM)~\cite{bellec2012pipeline}, was integrated
in CBRAIN using Boutiques. The integration uses the parallelization
mechanism described in Section~\ref{sec:parallelization} so that even
the invocations processing a single subject can be parallelized. It also
allows CBRAIN to leverage the efficient agent model used in PSOM, as
described in~\cite{GLAT-16}. The integration required some work in
PSOM to facilitate its invocation as a non-interactive command-line
application. The resulting CBRAIN plugin is available at
\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It could in
principle be used in other platforms that support the parallelization
property, although this has not been tested.

@@

\paragraph{Other applications.} Other neuroinformatics applications were ported to CBRAIN using Boutiques:
FSL tools and ICA AROMA are available at
\url{https://github.com/aces/cbrain-plugins-neuro} and some pipelines
of the Human Connectome Project (HCP) are at
\url{https://github.com/big-data-lab-team/cbrain-plugins-hcp}.

@@

\begin{itemize}
\item Security: containers may be updated after being integrated in the platforms.
\item The manifest specification is getting increasingly complex, but
  this complexity is required to describe real-world
  applications. When something cannot be described using the manifest,
  it is always possible to change the command line format by writing a
  wrapper script for the application. However, when containers are
  used, the wrapper script needs to be added to the container, which
  has some overhead (in particular, the containers provided by the
  developers cannot be used directly).
\item Testing is critical.
\item Workflow engines can be platforms (they can run Boutiques applications), workflows can be applications (they can be described in Boutiques).
\item With Boutiques, developers can port their application once and execute it in several platforms. It removes the technological dependency to a particular platform and facilitates migration. 
\end{itemize}

The \boutiques specification adopts a bottom-up approach where new
features are progressively added when applications or platforms
require them. It is currently supported by \todo{X} platforms and  \todo{Y} applications
were described.

 Although the motivating use cases
were taken from neuroinformatics, our primary field of interest,
nothing prevents the system from being used in a variety of other
fields.

@@

Boutiques intentionally does not support the definition of complete
workflows from individual applications. Indeed, we believe that
workflow engines are specific applications that need complex languages to
implement the particular functions that they target. Efforts to
specify common workflow languages include the Common Workflow Language
(CWL)~\cite{cwl} and the Interoperable Workflow Intermediate
Representation (IWIR)~\cite{plankensteiner2011iwir}.
